---
title: "Problem set 2: Regression"
author: "Hunter Giles"
date: "September 1, 2022"
output: 
  word_document: 
    toc: yes
  pdf_document: 
    latex_engine: xelatex
    toc: yes
  html_document: 
    toc: yes
---

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)
library(modelsummary)


# Load penguins data
penguins <- read_csv("data/penguins.csv")
```

# Task 1: Penguins

Between 2007 and 2009, researchers collected data on penguins in three islands in the Palmer Archipelago in Antarctica: Biscoe, Dream, and Torgersen. The `penguins` dataset has data for 342 penguins from 3 different species: Chinstrap, Gentoo, and Adélie. It includes the following variables:

-   `species`: The penguin's species (Chinstrap, Gentoo, and Adélie)
-   `island`: The island where the penguin lives (Biscoe, Dream, and Torgersen)
-   `bill_length_mm`: The length of the penguin's bill, in millimeters (distance from the penguin's face to the tip of the bill)
-   `bill_depth_mm`: The depth of the penguin's bill, in millimeters (height of the bill; distance from the bottom of the bill to the top of the bill)
-   `flipper_length_mm`: The length of the penguin's flippers, in millimeters
-   `body_mass_g`: The weight of the penguin, in grams
-   `sex`: The sex of the penguin
-   `year`: The year the observation was made

## Exploratory analysis

What is the relationship between penguin weight and bill depth? This plot shows some initial trends:

```{r plot-penguin-weight-depth}
ggplot(data = penguins, 
       aes(x = bill_depth_mm, y = body_mass_g)) +
  geom_point()
```

Make a new plot that colors these points by species. What can you tell about the relationship between bill depth and penguin weight?

```{r plot-penguin-weight-depth-by-species}
ggplot(data = penguins, 
       aes(x = bill_depth_mm, y = body_mass_g, color = species)) +
  geom_point()
```

*There is a positive relationship between bill depth and body mass.*

Add a `geom_smooth()` layer to the plot and make sure it uses a straight line (hint: include `method="lm"` in the function). What does this tell you about the relationship between bill depth and body mass?

```{r plot-penguin-weight-depth-by-species-with-lines}
ggplot(data = penguins, 
       aes(x = bill_depth_mm, y = body_mass_g, color = species)) +
  geom_point() +
  geom_smooth(method = lm)
```

*The lines show the average increase in body mass as bill depth increases for a given species.*

Change the plot so that there's a single line for all the points instead of one line per species. How does the slope of this single line differ from the slopes of the species specific lines? ***Why??***

```{r plot-penguin-weight-depth-by-species-with-one-line}
ggplot(data = penguins, 
       aes(x = bill_depth_mm, y = body_mass_g)) +
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = lm)
```

*This lines shows that there is a negative relationship between bill depth and body mass. From this visual we can infer that species biases the relationship.*

What is the relationship between flipper length and body mass? Make another plot with `flipper_length_mm` on the x-axis, `body_mass_g` on the y-axis, and points colored by `species`.

```{r}
ggplot(data = penguins, 
       aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() 
```

*There is a positive relationship between flipper length and body mass.*

Facet the plot by island (`island`)

```{r}
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  facet_wrap(vars(island))
```

Tell a story about the relationship between flipper length and weight in these three penguin species.

*There is a positive correlation between body mass and flipper length for all species. On average the Gentoo species weighs more and has bigger flippers, while the Adelie and Chinstrap are smaller birds.*

Tell a story about the distribution of penguins across the three islands.

*The Adelie seems to be the most versatile bird, they live on all three islands, while Chinstrap and Gentoo are restricted to their respective habitats.*

## Models

### Predicting weight with bill depth

Does bill depth predict penguin weight?

```{r model-depth-weight}
model_depth_weight <- lm(body_mass_g ~ bill_depth_mm,
                         data = penguins)

tidy(model_depth_weight, conf.int = TRUE)
```

```{r model-details-depth-weight}
glance(model_depth_weight)
```

INTERPRET THE COEFFICIENTS AND RESULTS HERE. What happens as bills get taller? Is the association statistically significant? How confident are you about these results? (Hint: look at the $R^2$)

*On average, as bill_depth increases by 1 mm, body_mass significantly decreases by 191 grams (p\<.001). However, the simple regression only explains 22% of the variation (R\^2 = .22), so we can infer that the model may suffer from an omitted variable problem.*

### Predicting weight with bill depth and flipper length

RUN A MODEL that predicts weight with bill depth and flipper length (i.e. body_mass_g \~ bill_depth_mm + flipper_length_mm)

```{r model-weight-depth-flipper}
reg1 <- lm(body_mass_g ~ bill_depth_mm + flipper_length_mm, data = penguins)
```

```{r model-details-weight-depth-flipper}
glance(reg1)
```

INTERPRET THESE RESULTS. Did the size of the bill depth coefficient change after controlling for flipper length?

*On average, as bill_depth increases by 1mm, body_mass increases by 22.6 grams at a .10 significance level. 1 mm increase in flipper_length increases body_mass by 51.5 grams (p\<.01). The previous model's omitted variable, flipper_length, biased down the bill_depth coefficient.*

### Predicting weight with bill depth, flipper length, and species

RUN A MODEL that predicts weight with bill depth, flipper length, and species.

```{r model-weight-depth-flipper-species}
reg2 <- lm(body_mass_g ~ bill_depth_mm + flipper_length_mm + species, data = penguins)
```

```{r model-details-weight-depth-flipper-species}
glance(reg2)
```

INTERPRET THESE RESULTS. What do the species coefficients mean? Did the bill depth coefficient change after controlling for both flipper length and species?

*In this model the Adelie species is used as a reference, so, on average, the Chinstrap is 131.97 grams lighter than the Adelie and the Gentoo is 1288 grams heavier. From the above results, we can see that species biases the bill_depth coefficient down when omitted.*

## All models at the same time

```{r all-penguin-models}
# Right now there's only one model here. Add the others from above (whatever you
# called them) like so: 
modelsummary(list(model_depth_weight, reg1, reg2))
```

------------------------------------------------------------------------

# Task 2: Food access and mortality

```{r load-food-mortality-data}
# Make sure you look at this dataset by clicking on its name in the Environment
# panel in RStudio. Sort some of the different columns and look around to get a
# feel for what's in the data
food_health <- read_csv("data/food_health_politics.csv") 
```

We're interested in looking at the relationships between food access, mortality, and politics. Do do this, we look at data from three different sources:

-   The USDA's [Food Environment Atlas](https://www.ers.usda.gov/data-products/food-environment-atlas/documentation/)
-   The CDC's ["Compressed Mortality File 1999-2015 Series 20 No. 2U, 2016"](http://wonder.cdc.gov/cmf-icd10.html)
-   2016 election results (found all over the internet)

Each row in the dataset is a US county. The main outcome we care about is `mortality_rate`, or the number of deaths per 100,000 people in a county between 2013-2015. Other interesting variables in the dataset include:

-   `pct_low_access_pop`: Percent of the county's population with low access to food
-   `pct_children_low_access`: Percent of the county's children with low access to food
-   `grocery_stores_per_1000`: Number of grocery stores in a county (per 1,000 residents)
-   `snap_stores_per_1000`: Number of stores that accept SNAP (food stamps) in a county (per 1,000 residents)
-   `fastfood_per_1000`: Number of fast food stores in a county (per 1,000 residents)
-   `per_dem_2012`: Percent of the county that voted for Obama in 2012
-   `per_dem_2016`: Percent of the county that voted for Clinton in 2016

## Exploratory analysis

### How related are mortality rate and access to food?

```{r cor-mortality-food}
# Notice how this is a little different from what was in the complete example
# with SAT scores. It's not possible to calculate correlations when there is
# missing data. The `use = "complete.obs"` argument here tells R to ignore any
# rows where either mortality_rate or pct_low_access_pop is missing
cor(food_health$mortality_rate, food_health$pct_low_access_pop,
    use = "complete.obs")
```

SAY SOMETHING HERE. This is backwards from what you might expect, since it trends downward (i.e. the mortality rate is lower in counties with a greater proportion of the population with low access to food). Why might that be? Is there really a relationship?

*Intuitively, if the population has less access to food, then the mortality rate should increase. The negative relationship above is relatively weak, so there is likely another variable influencing this relationship.*

```{r plot-mortality-food, warning=FALSE}
# Use warning=FALSE in the chunk options to remove the warning about missing data
ggplot(food_health, aes(x = pct_low_access_pop, y = mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "% of county with low access to food", 
       y = "Mortality rate (per 100,000 residents)")
```

### How related are mortality rate and the prevalence of fast food restaurants?

```{r cor-mortality-fastfood}
cor(food_health$mortality_rate, food_health$fastfood_per_1000,
    use = "complete.obs")
```

SAY SOMETHING HERE

*Mortality rate and number of fastfood restaurants have a weak negative correlation. This makes since, while fastfood does help hunger, it also increases the risk of cardiovascular disease. The two opposing effects cancel out.*

```{r plot-mortality-fastfood}
ggplot(food_health, aes(x = fastfood_per_1000, y = mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Number of fast food stores in a county (per 1,000 residents)", 
       y = "Mortality rate (per 100,000 residents)")
```

### How related are mortality rate and the prevalence of SNAP stores per 1,000 residents?

```{r cor-mortality-snap}
cor(food_health$mortality_rate, food_health$snap_stores_per_1000, 
    use = "complete.obs")
```

SAY SOMETHING HERE

*Mortality and number of store that accept snap have a moderate positive correlation. Counties that accept SNAP are likely to be poor which increases mortality.*

```{r plot-mortality-snap}
ggplot(food_health, aes(x = snap_stores_per_1000, y = mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Number of stores that accept SNAP (per 1,000 residents)", 
       y = "Mortality rate (per 100,000 residents)")
```

### How related are mortality rate and the percent of the county that voted for Democrats in 2016?

```{r cor-mortality-2016}
cor(food_health$mortality_rate, food_health$votes_dem_2016, 
    use = "complete.obs")
```

SAY SOMETHING HERE

*There is a weak negative correlation between mortality rates and the number of democratic votes in 2016.*

```{r plot-mortality-2016}
ggplot(food_health, aes(x = votes_dem_2016, y = mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Percent of the county that voted for Clinton in 2016", 
       y = "Mortality rate (per 100,000 residents)")
```

## Models

### Does access to food predict mortality?

SAY SOMETHING HERE

*Intuitively, low access to food should have a causal relationship with mortality. However, the above analysis only shows correlation.*

```{r model-mortality-food}
model_mortality_food <- lm(mortality_rate ~ pct_low_access_pop,
                           data = food_health)

tidy(model_mortality_food, conf.int = TRUE)
# View(select(food_health, mortality_rate, pct_low_access_pop))
```

```{r model-details-mortality-food}
glance(model_mortality_food)
```

INTERPRET THE COEFFICIENTS AND RESULTS HERE. What happens as the percent of low access to food goes up by 1%? Is that significant? Again, this is backwards from what you'd expect---as the percent of low access goes up, mortality drops. Why might that be? How much do you trust this finding? (Hint: look at the R2 value)

*On average, as pct_low_access_pop increases by 1%, the county's mortality rate drops by 1.24 (p\<.001). The results seem backwards. This simple regression accounts for only 2.6% of the variation in mortality, meaning there are other variables biasing the pct_low_access_pop coefficient.*

### Do more SNAP stores per person predict mortality?

```{r model-mortality-snap}
reg5 <- lm(data=food_health, mortality_rate ~ snap_stores_per_1000)
tidy(reg5, conf.int = TRUE)
```

```{r model-details-mortality-snap}
glance(reg5)
#View(select(food_health, mortality_rate, snap_stores_per_1000))
```

SAY SOMETHING HERE. What happens as the proportion of SNAP stores goes up? Do you trust this number more or less than low access to food?'

*As snap_stores_per_1000 increases by 1, the mortality rate rises by 174. snap_stores_per_1000 is a much better predictor than pct_low_access_pop. We can infer this by looking at this model's R\^2 score (R\^2 = .19). Intuitively, this makes sense. SNAP stores ate primarily in lower class locations, which have higher criminal activity, lower quality education, etc.*

### Do election results and access to food and SNAP stores predict mortality?

RUN A MODEL THAT PREDICTS MORTALITY WITH A BUNCH OF COVARIATES (i.e. mortality_rate \~ pct_low_access_pop + snap_stores_per_1000 + per_dem_2016 + anything else you want to throw in)

```{r model-mortality-lots-of-things}
reg6 <- lm(mortality_rate ~ pct_low_access_pop + snap_stores_per_1000 + per_dem_2016 + grocery_stores_per_1000, data = food_health)
tidy(reg6, conf.int = TRUE)
```

```{r model-details-mortality-lots-of-things}
glance(reg6)
#View(select(food_health, mortality_rate, pct_low_access_pop, snap_stores_per_1000, per_dem_2016, grocery_stores_per_1000 ))
```

SAY SOMETHING HERE. Interpret the different coefficients. How predictive is this model (i.e. what's the R2)? Do you believe this model?

*On Average, as pct_low_access_pop increases by 1%, mortality rate decreases by .86. As snap_stores_per_1000 increases by 1, mortality rate increases by 217. As per_dem_2016 increases by 1%, mortality rate decreases by 1.8. As grocery_stores_per_1000 increases by 1, mortality rate decreases by 220.89. This model accounts for 34.13% of the variation in mortality rate. The the models robustness has greatly increases from the simple regressions, so I am more likely to believe this one.*

### Mortality, contolling for state differences

RUN A MODEL with some number of plausible independent/explanatory variables. Include `state` as one of them

```{r model-mortality-state}
# Add other explanatory variables here
model_with_state <- lm(mortality_rate ~ pct_low_access_pop + state + snap_stores_per_1000 ,
                       data = food_health)

# This table is 50+ rows long! While it might be interesting to see changes in
# intercept in relation to Alaska (the omitted state here), like how Alabama's
# mortality rate is 137 higher than Alaska's while DC's is 84 lower, it's not
# super helpful. Controlling for state does capture some of the state-specific
# reasons for varying mortality though, so it's good to include. We just don't
# really need to see all those coefficients. To remove them from this table of
# results, filter them out. The "!" in R means "not", so here we're only looking
# at rows that don't start with "state"
tidy(model_with_state, conf.int = TRUE) %>% 
  filter(!str_starts(term, "state"))
```

```{r model-state-mortality-lots-of-things}
glance(model_with_state)
```

SAY SOMETHING ABOUT THIS MODEL

*By including state fixed effects, we are about to account for 48% of the variation in mortality rate.*

## All models at the same time

PUT ALL THE MODEL RESULTS IN THE SAME SIDE-BY-SIDE TABLE HERE

```{r everything-together}
# Right now there are only two models here. Add the others from above (whatever 
# you called them) like so: 
# modelsummary(list(model_mortality_food, some_other_model, yet_another_model, etc))

# Also, by default, modelsummary will include all the state coefficients, which
# we don't want. We can omit specific coefficients with the `coef_omit`
# argument. The ^ character means it'll omit coefficients that *start with*
# "state". Without ^, it would omit any coefficient where the characters "state"
# appeared anywhere in the name, which might be too greedy

modelsummary(list(model_mortality_food, model_with_state, reg5, reg6),
             coef_omit = "^state")
```
