---
title: "Problem Set 4"
author: "W. Hunter Giles"
format: 
    pdf:
      code-line-numbers: true
      code-block-bg: true
    gfm: 
      toc: true
      
editor: visual
---

------------------------------------------------------------------------

In 1996, Nada Eissa and Jeffrey B. Liebman [published a now-classic study on the effect of the Earned Income Tax Credit (EITC) on employment](http://darp.lse.ac.uk/papersdb/Eissa-Liebman_(QJE96).pdf). The EITC is a special tax credit for low income workers that changes depending on (1) how much a family earns (the lowest earners and highest earners don't receive a huge credit, as the amount received phases in and out), and (2) the number of children a family has (more kids = higher credit). See [this brief explanation](https://www.cbpp.org/research/federal-tax/policy-basics-the-earned-income-tax-credit) for an interactive summary of how the EITC works.

Eissa and Liebman's study looked at the effects of the EITC on women's employment and wages after it was initially substantially expanded in 1986. The credit was expanded substantially again in 1993. For this problem set, you'll measure the causal effect of this 1993 expansion on the employment levels and annual income for women.

A family must have children in order to quality for the EITC, which means the presence of 1 or more kids in a family assigns low-income families to the EITC program (or "treatment"). We have annual data on earnings from 1991--1996, and because the expansion of EITC occurred in 1993, we also have data both before and after the expansion. This treatment/control before/after situation allows us to use a difference-in-differences approach to identify the causal effect of the EITC.

The dataset I've provided (`eitc.dta`) is a Stata data file containing more than 13,000 observations. This is non-experimental data---the data comes from the US Census's Current Population Survey (CPS) and includes all women in the CPS sample between the ages of 20--54 with less than a high school education between 1991--1996. There are 11 variables:

-   `state`: The woman's state of residence. The numbers are Census/CPS state numbers: <http://unionstats.gsu.edu/State_Code.htm>
-   `year`: The tax year
-   `urate`: The unemployment rate in the woman's state of residence
-   `children`: The number of children the woman has
-   `nonwhite`: Binary variable indicating if the woman is not white (1 = Hispanic/Black)
-   `finc`: The woman's family income in 1997 dollars
-   `earn`: The woman's personal income in 1997 dollars
-   `age`: The woman's age
-   `ed`: The number of years of education the woman has
-   `unearn`: The woman's family income minus her personal income, in *thousands* of 1997 dollars

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)  # For ggplot, %>%, mutate, filter, group_by, and friends
library(haven)      # For loading data from Stata
library(broom)      # For showing models as data frames

# This turns off this message that appears whenever you use summarize():
# `summarise()` ungrouping output (override with `.groups` argument)
options(dplyr.summarise.inform = FALSE)

# Load EITC data
eitc <- read_stata("../data/eitc.dta") %>% 
  # case_when() is a fancy version of ifelse() that takes multiple conditions
  # and outcomes. Here, we make a new variable named children_cat(egorical) 
  # with three different levels: 0, 1, and 2+
  mutate(children_cat = case_when(
    children == 0 ~ "0",
    children == 1 ~ "1",
    children >= 2 ~ "2+"
  ))
```

# 1. Exploratory data analysis

## Work

```{r}
# Work
eitc %>% 
  group_by(children_cat) %>% 
  summarize(avg_work = mean(work))

# stat_summary() here is a little different from the geom_*() layers you've seen
# in the past. stat_summary() takes a function (here mean_se()) and runs it on
# each of the children_cat groups to get the average and standard error. It then
# plots those with geom_pointrange. The fun.args part of this lets us pass an
# argument to mean_se() so that we can multiply the standard error by 1.96,
# giving us the 95% confidence interval
ggplot(eitc, aes(x = children_cat, y = work)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96)) +
  labs(title = "Average Work per Child")
```

## Family income

```{r}
eitc %>% 
  group_by(children_cat) %>%
  summarise(avg_inc = mean(finc))

ggplot(eitc, aes(x = children_cat, y = finc)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96)) +
  labs(title = "Average Family Income per Child")
  
```

## Earnings

```{r}
eitc %>% 
  group_by(children_cat) %>%
  summarise(avg_inc = mean(earn))

ggplot(eitc, aes(x = children_cat, y = earn)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96)) +
  labs(title = "Average Personal Income per Child")

```

## Race

```{r}
eitc %>% 
  group_by(children_cat) %>%
  summarise(avg_inc = mean(nonwhite))

ggplot(eitc, aes(x = children_cat, y = nonwhite)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96)) +
  labs(title = "Average Non-White per Child")

```

## Education

```{r}
eitc %>% 
  group_by(children_cat) %>%
  summarise(avg_inc = mean(ed))

ggplot(eitc, aes(x = children_cat, y = ed)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96)) +
  labs(title = "Average Education per Child")
```

## Age

```{r}
eitc %>% 
  group_by(children_cat) %>%
  summarise(avg_inc = mean(age))

ggplot(eitc, aes(x = children_cat, y = age)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96)) +
  labs(title = "Average Age per Child")

```

## General summary

The number of children has a strictly monotonic relationship with `work`, `finc`, `earn`, `nonwhite`, `ed`, and `age` (be aware that children is binned for \>=2). These relationships are not yet proven to be causal, so the above results should be treated as correlation. As probability of employment (`work`) increases, the number of children decrease. As family income (`finc`) increases, the number of children decrease. As personal income (`earn`) increase the number of children decreases. As education (`ed`) increases, the number of children increases. As families become more non-white (`non-white`), the number of children decrease. As age (`age`) increases, then number of children decrease.

# 2. Create treatment variables

```{r}

# dummy for has any children
eitc <- eitc %>%
  mutate(any_kids = ifelse(children > 0,1,0)) 

# dummy for after 1993
eitc <- eitc %>% 
  mutate(after_1993 = year > 1993) 

# View(select(eitc, any_kids, children, after_1993, year))
```

# 3. Check pre- and post-treatment trends

```{r}
eitc_by_year_kids <- eitc %>% group_by(any_kids, year) %>% 
  summarise(avg_work = mean(work))
print(eitc_by_year_kids)
```

### Plot for Parallel Trends

```{r}
ggplot(data = eitc_by_year_kids) +
  geom_point(mapping = aes(x = year, y = avg_work, color=factor(any_kids))) +
  geom_line(mapping = aes(x = year, y = avg_work, group=any_kids, color=factor(any_kids))) +
  geom_vline(xintercept = 1994) +
  labs(title = "Probability of Working (1991 - 1996)", x = "Years", y = "Probability of Working", color = " Any Kids") 
```

Parallel trends can be assumed for the pre-treatment period.

# 4. Difference-in-difference by hand-ish

```{r}
a <- eitc %>% group_by(any_kids, after_1993) %>%
  summarise(avg_work = mean(work))
a <- spread(a, key = after_1993, value = avg_work)
a
```

```{r}
mutate(a, first_difference_time = `TRUE` - `FALSE`)
data.frame(t(a)[2:3,]) %>% mutate(first_difference_kids = X1 - X2)
```

```{r}
0.044799623 - -0.002073509
0.12949788 - 0.08262475
```

|                    | Before 1993 | After 1993 | Difference   |
|--------------------|-------------|------------|--------------|
| Women with no kids | 0.5754597   | 0.5733862  | -0.002073509 |
| Women with kids    | 0.4459619   | 0.4907615  | 0.044799623  |
| Difference         | 0.12949788  | 0.08262475 | 0.04687313   |

From the table above, we can see that women without kids works .2% after the EITC tax, and women with kids worked 4.4% more. In the pre-treatment period, 12.9% of women without kids worked more than women with kids. After the treatment, 8% of women without kids worked more than women with kids. The difference and difference shows that the EITC increases the number of women with kids that worked by 4.6%.

# 5. Difference-in-difference with regression

```{r}
lm_model <- lm(work ~ (any_kids * after_1993), data = eitc)
tidy(lm_model)
```

The diff-and-diff regression yields the same results as the table in part 4. The advantage of the regression, is we can can see the standard error, t statistics, and p values. Now we can say that the EITC policy significantly (p\<.01) increase the number of women with kids that work by 4.6%.

# 6. Difference-in-difference with regression and controls

```{r}
# Make new dataset with columns for age squared and education squared
# Regression model with demographic controls here
eitc <- eitc %>% 
  mutate(age_sq = age*age, ed_sq = ed*ed)

lm_model_controls <- lm(work ~ (any_kids * after_1993) + age + age_sq + ed + ed_sq + unearn + nonwhite, data = eitc)
tidy(lm_model_controls)
```

This model include variables that could be confounding the main difference and difference (DD) interaction. This includes two squared variables, `age_sq` and `ed_eq`, that account for non linear returns in percent worked explained by age and education. With all controls, we can see that the DD interaction term's coefficient went up, meaning the leaving the co-variates out of the model, biased the interaction down. The treatment effect is now interpreted as significantly (p \< .01) increasing the number of women with kids that worked by 5.8%.

# 7. Varying treatment effects

```{r}
eitc <- eitc %>% mutate(one_kid = ifelse(children == 1, 1, 0))
eitc <- eitc %>% mutate(two_plus_kid = ifelse(children > 1, 1, 0))
```

```{r}
lm_model_differentiated_kids <- lm(work ~ (one_kid * after_1993) + (two_plus_kid * after_1993), data = eitc)

tidy(lm_model_differentiated_kids)
```

Form the regression above, we can see that the treatment has a stronger effect on women with 2 or more kids. One intuitive reasoning for this is women with more kids receive more funding. This gives them a greater incentive to work than women with one kid.

# 8. Check parallel trends with fake treatment

```{r}
eitc_fake_treatment <- eitc %>% 
  filter(year < 1994) %>% 
  mutate(after_1991 = year >= 1992)
# Make new dataset that only includes rows less than 1994 (with filter), and add
# a new binary indicator variable for after_1991

# Run simple regression with interaction term any_kids * after_1991
lm_fake_model <- lm(work ~ (any_kids * after_1991), data = eitc_fake_treatment)
tidy(lm_fake_model)
```

The effect is both small and insignificant. The fake treatment procedure serves as a robustness check. We can now me more certain that the treatment caused the change in work, and the DD model is measuring the intended effect.
